Generalized Bradley-Terry models

### 2025-03-20 inst/jss_hyper3_real.Rnw
JSS bibliography style [capialization]

One generalization of the Bradley-Terry model is due
to~\citet{luce1959}, in which the probability of competitor~$i$
winning in a field of~$j=1,\ldots, n$ is $\frac{p_i}{p_1+\cdots
  +p_n}$.  It is then natural to quantify some ephemeral enhancement
to competitor~$i$ using a multiplicative factor $\lambda$.  Then
$\Prob(\mbox{$i$ wins})=\frac{\lambda p_i}{p_1+\cdots+\lambda
  p_i+\cdots+p_n}$.  Examples of such enhancements include the
home-ground advantage in football and rugby, or playing white in
chess~\citep{hankin2020}.  Note that $\lambda < 1$ would correspond to
a decrease in competitive ability.

Further, one might follow~\cite{hankin2017_rnw} and consider teams of
players.  In doubles tennis, for example, $\left\lbrace
p_1,p_2\right\rbrace$ might play $\left\lbrace p_3,p_4\right\rbrace$
and win with probability
$\frac{p_1+p_2}{p_1+p_2+p_3+p_4}$~\citep{hankin2010}.  It is natural
to combine this idea with that of a multiplicative factor.  For
example, suppose $p_1$ has just learned a new special technique, and
$p_4$ has just been given a performance-enhancing drug.  If so, we
might have likelihood function
%
\begin{equation}\label{like1234}
  \mathcal{L}{\left(p_1,p_2,p_3,p_4;\lambda,\mu\right)}\propto
  \frac{
    \left(\lambda p_1 + p_2\right)^{w_{12}}\cdot
    \left(p_3 + \mu p_4\right)^{w_{34}}
    }{
    \left(\lambda p_1+p_2+p_3+\mu p_4\right)^{w_{12} + w_{34}}
    }
\end{equation}
%
where $w_{12}$ and $w_{34}$ are the number of wins by each team
respectively.  Then $\lambda$ would represent the effectiveness of
$p_1$'s special technique, and $\mu$ the enhancement afforded to $p_4$
by the drug; values $>1$ would correspond to a positive effect.  In
this paper I generalize the \pkg{hyper2} R
package~\citep{hankin2017_rnw}, available on CRAN~\citep{rcore2025},
for Bradley-Terry likelihoods to accommodate this class of
multiplicative factors with new \proglang{S3} \class{hyper3} and
associated methods.

The most general likelihood function would be
proportional to
%
\begin{equation}\label{mostgeneral}
  \prod_{i=1}^N
  \sum_{j=1}^n
 \left(\lambda_{ij}p_j\right)^{w_i}
\end{equation}
%
where the $p_j$ are traditional Bradley-Terry strengths, $p_j\geqslant
0$, $\sum p_j=1$, and the $\lambda_{ij}\geqslant 0$ are multiplicative
factors that account for ephemeral or temporary changes in competitive
ability.  The data $w_1,\ldots w_N$ correspond to the number of
``wins'' and number of matches, as per $w_{12},w_{34}$ in
equation~\ref{like1234}.  In practice, almost all the $\lambda_{ij}$
are zero.  To fix ideas, consider the data presented in
Table~\ref{oldfirm}.  Ignoring draws for the moment, we have a
likelihood proportional to
%
\begin{table}[t!]
\centering
\begin{tabular}{@{} *{5}{c} @{}}
\headercell{\small{Plays at home,}\\ \small{wins}} &
\multicolumn{4}{c@{}}{\small{Plays away, loses}}\\
\cmidrule(l){2-5}
&   Rangers     &    Celtic & Falkirk & Livingston\\
\midrule
   Rangers      &   NA &    81  &    69   &      15\\
   Celtic       &   76 &    NA  &    61   &      15\\
   Falkirk      &   17 &    17  &    NA   &       6\\
   Livingston   &    2 &     2  &     3   &      NA\\
\end{tabular}
\\
\rule{0mm}{16mm}
\\
\begin{tabular}{@{} *{5}{c} @{}}
\headercell{\small{Plays at home,}\\ \small{draws}} &
\multicolumn{4}{c@{}}{\small{Plays away, draws}}\\
\cmidrule(l){2-5}
&  Rangers     &    Celtic & Falkirk & Livingston\\
\midrule
   Rangers     &       NA  &   44 &     14   &       2\\
   Celtic      &       47  &   NA &     10   &       3\\
   Falkirk     &       12  &   19 &     NA   &       4\\
   Livingston  &       4   &   4  &     4    &      NA\\
   \end{tabular}
\\
\rule{0mm}{16mm}
\\
\begin{tabular}{@{} *{5}{c} @{}}
\headercell{\small{Plays at home,}\\ \small{loses}} &
\multicolumn{4}{c@{}}{\small{Plays away, wins}}\\
\cmidrule(l){2-5}
&  Rangers     &    Celtic & Falkirk & Livingston\\
\midrule
   Rangers     &   NA  &  42  &    2  &       0\\
   Celtic      &   46  &  NA  &    8  &       0\\
   Falkirk     &   57  &  43  &   NA  &       8\\
   Livingston  &   12  &  11  &    9  &      NA\\
  \end{tabular}
\caption{The \label{oldfirm} dominance of the Old Firm: 769 matches
  between Rangers, Celtic, Falkirk and Livingston arranged by home
  wins, draws, and home losses.  Data from {\tt stats.football.co.uk}}
\end{table}
%
\begin{equation}\label{firstfewRCFL}
  \left(\frac{\lambda p_R}{\lambda p_R + p_C}\right)^{81}\cdot
  \left(\frac{\lambda p_R}{\lambda p_R + p_F}\right)^{69}\cdots
  \left(\frac{\lambda p_L}{\lambda p_L + p_F}\right)^{3}\cdots
  \left(\frac{p_C}{p_C + \lambda p_R}\right)^{42}\cdots
    \left(\frac{p_F}{p_F + \lambda p_L}\right)^{9}
\end{equation}
%
where the home-ground advantage is represented by~$\lambda$.  In
package idiom, a term such as $\lambda p_R + p_C$ is represented as a
named numeric vector.  For example, if the home strength were $\lambda
= 1.8$ then the first bracket would be {\tt c(Rangers=1.8, Celtic=1)}
with an associated power of $-81$.  The computational methodology is a
generalization of that used in~\cite{hankin2024_hyper3} in which the
multiplicative factors correspond to number of clones.  However,
in~\cite{hankin2024_hyper3} the factors are known non-negative
integers, and here they are, in general, unknown non-negative real
numbers.  The \class{hyper3} class enforces \pkg{disordR}
discipline~\citep{hankin2022_disordR}.

@@

About 22\% of matches in Table~\ref{oldfirm} end in a draw and {\tt
  hyper2} formalism may be used to assess this.  Consider the
likelihood function described in Table~\ref{windrawloss}; this has a
number of desirable features not present in the reified entity
approach of~\citet{hankin2020}.  Firstly, if we consider two players
of identical strengths and nullify the home ground effect (that is,
$\lambda=1$) then the probability of a draw is $\frac{2D}{2+2D}$; the
odds of a draw is simply $D$, regardless of team strength.  Secondly,
considering only won (that is, not score-drawn) games, the probability
of a home win is $\frac{\lambda}{1+\lambda}$, that is, the odds of a
home win is simply $\lambda$, independently of the player
strength. Such arguments give direct operational significance to
$\lambda$ and $D$.

>>>>

One generalization of the Bradley-Terry model is due
to~\citet{luce1959}, in which the probability of competitor~$i$
winning in a field of~$j=1,\ldots, n$ is $\frac{p_i}{p_1+\cdots
  +p_n}$.  It is then natural to quantify some ephemeral enhancement
to competitor~$i$ using a multiplicative factor $\lambda$.  Then
$\Prob(\mbox{$i$ wins})=\frac{\lambda p_i}{p_1+\cdots+\lambda
  p_i+\cdots+p_n}$.  Examples of such enhancements include the
home-ground advantage in football and rugby, or playing white in
chess~\citep{hankin2020_jss}.  Note that $\lambda < 1$ would correspond to
a decrease in competitive ability.

Further, one might follow~\cite{hankin2017_rnw_jss} and consider teams of
players.  In doubles tennis, for example, $\left\lbrace
p_1,p_2\right\rbrace$ might play $\left\lbrace p_3,p_4\right\rbrace$
and win with probability
$\frac{p_1+p_2}{p_1+p_2+p_3+p_4}$~\citep{hankin2010}.  It is natural
to combine this idea with that of a multiplicative factor.  For
example, suppose $p_1$ has just learned a new special technique, and
$p_4$ has just been given a performance-enhancing drug.  If so, we
might have likelihood function
%
\begin{equation}\label{like1234}
  \mathcal{L}{\left(p_1,p_2,p_3,p_4;\lambda,\mu\right)}\propto
  \frac{
    \left(\lambda p_1 + p_2\right)^{w_{12}}\cdot
    \left(p_3 + \mu p_4\right)^{w_{34}}
    }{
    \left(\lambda p_1+p_2+p_3+\mu p_4\right)^{w_{12} + w_{34}}
    }
\end{equation}
%
where $w_{12}$ and $w_{34}$ are the number of wins by each team
respectively.  Then $\lambda$ would represent the effectiveness of
$p_1$'s special technique, and $\mu$ the enhancement afforded to $p_4$
by the drug; values $>1$ would correspond to a positive effect.  In
this paper I generalize the \pkg{hyper2} R
package~\citep{hankin2017_rnw_jss}, available on CRAN~\citep{rcore2025},
for Bradley-Terry likelihoods to accommodate this class of
multiplicative factors with new \proglang{S3} \class{hyper3} and
associated methods.

The most general likelihood function would be
proportional to
%
\begin{equation}\label{mostgeneral}
  \prod_{i=1}^N
  \sum_{j=1}^n
 \left(\lambda_{ij}p_j\right)^{w_i}
\end{equation}
%
where the $p_j$ are traditional Bradley-Terry strengths, $p_j\geqslant
0$, $\sum p_j=1$, and the $\lambda_{ij}\geqslant 0$ are multiplicative
factors that account for ephemeral or temporary changes in competitive
ability.  The data $w_1,\ldots w_N$ correspond to the number of
``wins'' and number of matches, as per $w_{12},w_{34}$ in
equation~\ref{like1234}.  In practice, almost all the $\lambda_{ij}$
are zero.  To fix ideas, consider the data presented in
Table~\ref{oldfirm}.  Ignoring draws for the moment, we have a
likelihood proportional to
%
\begin{table}[t!]
\centering
\begin{tabular}{@{} *{5}{c} @{}}
\headercell{\small{Plays at home,}\\ \small{wins}} &
\multicolumn{4}{c@{}}{\small{Plays away, loses}}\\
\cmidrule(l){2-5}
&   Rangers     &    Celtic & Falkirk & Livingston\\
\midrule
   Rangers      &   NA &    81  &    69   &      15\\
   Celtic       &   76 &    NA  &    61   &      15\\
   Falkirk      &   17 &    17  &    NA   &       6\\
   Livingston   &    2 &     2  &     3   &      NA\\
\end{tabular}
\\
\rule{0mm}{16mm}
\\
\begin{tabular}{@{} *{5}{c} @{}}
\headercell{\small{Plays at home,}\\ \small{draws}} &
\multicolumn{4}{c@{}}{\small{Plays away, draws}}\\
\cmidrule(l){2-5}
&  Rangers     &    Celtic & Falkirk & Livingston\\
\midrule
   Rangers     &       NA  &   44 &     14   &       2\\
   Celtic      &       47  &   NA &     10   &       3\\
   Falkirk     &       12  &   19 &     NA   &       4\\
   Livingston  &       4   &   4  &     4    &      NA\\
   \end{tabular}
\\
\rule{0mm}{16mm}
\\
\begin{tabular}{@{} *{5}{c} @{}}
\headercell{\small{Plays at home,}\\ \small{loses}} &
\multicolumn{4}{c@{}}{\small{Plays away, wins}}\\
\cmidrule(l){2-5}
&  Rangers     &    Celtic & Falkirk & Livingston\\
\midrule
   Rangers     &   NA  &  42  &    2  &       0\\
   Celtic      &   46  &  NA  &    8  &       0\\
   Falkirk     &   57  &  43  &   NA  &       8\\
   Livingston  &   12  &  11  &    9  &      NA\\
  \end{tabular}
\caption{The \label{oldfirm} dominance of the Old Firm: 769 matches
  between Rangers, Celtic, Falkirk and Livingston arranged by home
  wins, draws, and home losses.  Data from {\tt stats.football.co.uk}}
\end{table}
%
\begin{equation}\label{firstfewRCFL}
  \left(\frac{\lambda p_R}{\lambda p_R + p_C}\right)^{81}\cdot
  \left(\frac{\lambda p_R}{\lambda p_R + p_F}\right)^{69}\cdots
  \left(\frac{\lambda p_L}{\lambda p_L + p_F}\right)^{3}\cdots
  \left(\frac{p_C}{p_C + \lambda p_R}\right)^{42}\cdots
    \left(\frac{p_F}{p_F + \lambda p_L}\right)^{9}
\end{equation}
%
where the home-ground advantage is represented by~$\lambda$.  In
package idiom, a term such as $\lambda p_R + p_C$ is represented as a
named numeric vector.  For example, if the home strength were $\lambda
= 1.8$ then the first bracket would be {\tt c(Rangers=1.8, Celtic=1)}
with an associated power of $-81$.  The computational methodology is a
generalization of that used in~\cite{hankin2024_hyper3_jss} in which the
multiplicative factors correspond to number of clones.  However,
in~\cite{hankin2024_hyper3_jss} the factors are known non-negative
integers, and here they are, in general, unknown non-negative real
numbers.  The \class{hyper3} class enforces \pkg{disordR}
discipline~\citep{hankin2022_disordR_jss}.

@@

About 22\% of matches in Table~\ref{oldfirm} end in a draw and {\tt
  hyper2} formalism may be used to assess this.  Consider the
likelihood function described in Table~\ref{windrawloss}; this has a
number of desirable features not present in the reified entity
approach of~\citet{hankin2020_jss}.  Firstly, if we consider two players
of identical strengths and nullify the home ground effect (that is,
$\lambda=1$) then the probability of a draw is $\frac{2D}{2+2D}$; the
odds of a draw is simply $D$, regardless of team strength.  Secondly,
considering only won (that is, not score-drawn) games, the probability
of a home win is $\frac{\lambda}{1+\lambda}$, that is, the odds of a
home win is simply $\lambda$, independently of the player
strength. Such arguments give direct operational significance to
$\lambda$ and $D$.
